---
title: "k-NN Classification"
author: "Michael Roberto"
date: "2025-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
# Load all necessary libraries for the assignmnt
library(class)  
library(caret)  
library(dplyr)   
library(ggplot2) 
library(readr)   
```
```{r}
# Load dataset for assignment and remove the two unnecessary columns
data <- read.csv("UniversalBank.csv") %>% select(-ID, -ZIP.Code)
```
```{r}
# Convert categorical variable (Education) into three dummy variables
data <- data %>%
  mutate(
    Education_1 = ifelse(Education == 1, 1, 0),
    Education_2 = ifelse(Education == 2, 1, 0),
    Education_3 = ifelse(Education == 3, 1, 0)
  ) %>%
  select(-Education)
```
```{r}
#Normalize the numerical predictions
normalize <- function(x) { (x - min(x)) / (max(x) - min(x)) }
data[, 1:7] <- as.data.frame(lapply(data[, 1:7], normalize))
```
```{r}
# Split data into training set at 60% and validation set at 40%
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.6, list = FALSE)
trainData <- data[trainIndex, ]
validData <- data[-trainIndex, ]
```
```{r}
# Define the predictor and response variables
train_X <- trainData %>% select(-Personal.Loan)
train_Y <- trainData$Personal.Loan
valid_X <- validData %>% select(-Personal.Loan)
valid_Y <- validData$Personal.Loan
```
```{r}
# Apply the k-NN where k = 1 and evaluate outcome
knn_pred_k1 <- knn(train = train_X, test = valid_X, cl = train_Y, k = 1)
confusionMatrix(table(knn_pred_k1, valid_Y))
```
```{r}
# Find optimal k by evaluating accuracy where k = 1 and 20
k_values <- seq(1, 20, 2)
accuracy_results <- data.frame(k = k_values, Accuracy = NA)

for (i in seq_along(k_values)) {
  knn_pred <- knn(train_X, valid_X, train_Y, k = k_values[i])
  accuracy_results$Accuracy[i] <- sum(knn_pred == valid_Y) / length(valid_Y)
}
```
```{r}
# Plot accuracy of outcome versus k
ggplot(accuracy_results, aes(x = k, y = Accuracy)) +
  geom_line() + geom_point() +
  ggtitle("Optimal k") +
  xlab("Number of Neighbors (k)") +
  ylab("Accuracy")
```
```{r}
# Choose optimal k and evaluate outcomes
best_k <- accuracy_results$k[which.max(accuracy_results$Accuracy)]
knn_pred_best_k <- knn(train_X, valid_X, train_Y, k = best_k)
confusionMatrix(table(knn_pred_best_k, valid_Y))
```
```{r}
# Classification of a new customer using the best k option
new_customer <- data.frame(
  Age = normalize(40), Experience = normalize(10), Income = normalize(84),
  Family = normalize(2), CCAvg = normalize(2), Mortgage = normalize(0),
  Securities.Account = 0, CD.Account = 0, Online = 1, CreditCard = 1,
  Education_1 = 0, Education_2 = 1
)
```
# Repartition data between 50 training, 30 validation and 20 test
```{r}
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.5, list = FALSE)
trainData <- data[trainIndex, ]
remainingData <- data[-trainIndex, ]
validIndex <- createDataPartition(remainingData$Personal.Loan, p = 0.6, list = FALSE)
validData <- remainingData[validIndex, ]
testData <- remainingData[-validIndex, ]
```
```{r}
# Train and evaluate k-NN on new the 50/30/20 partitions
train_X <- trainData %>% select(-Personal.Loan)
train_Y <- trainData$Personal.Loan
valid_X <- validData %>% select(-Personal.Loan)
valid_Y <- validData$Personal.Loan
test_X <- testData %>% select(-Personal.Loan)
test_Y <- testData$Personal.Loan
```
```{r}
knn_test_pred <- knn(train_X, test_X, train_Y, k = best_k)
confusionMatrix(table(knn_test_pred, test_Y))
```
```{r}
#Comparison of confusion matric between train/test/validation sets
knn_train_pred <- knn(train_X, train_X, train_Y, k = best_k)
train_conf_matrix <- confusionMatrix(table(knn_train_pred, train_Y))

knn_valid_pred <- knn(train_X, valid_X, train_Y, k = best_k)
valid_conf_matrix <- confusionMatrix(table(knn_valid_pred, valid_Y))

knn_test_pred <- knn(train_X, test_X, train_Y, k = best_k)
test_conf_matrix <- confusionMatrix(table(knn_test_pred, test_Y))
```
```{r}
#Print of confusion matrix to compare outcomes
print("Train Confusion Matrix:"); print(train_conf_matrix)
print("Validation Confusion Matrix:"); print(valid_conf_matrix)
print("Test Confusion Matrix:"); print(test_conf_matrix)
```
#Assessment of project data:  The new customer is classifed as 0: no personal loan with the best k being 5-7.  Training accuracy will be high whereas validation accuracy will normalize k.  Test accuracy shows 'real' world performance but if test accuracy drops, the model will overfit.  Assuming validation and test accuract perform at similar levels, the model provides a good generalization of behavior.  
