---
title: "Hitters in R"
author: "Michael Roberto"
date: "2025-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
getwd()
```
```{r}
#load ISLR for dataset and readr/ggplots from Tidyverse
library(ISLR)
library(readr)
library(ggplot2)
```
```{r}
head(Hitters)
```
```{r}
tail(Hitters)
```
```{r}
summary(Hitters)
```
```{r}
dim(Hitters)
```
#Used functions like head, tail, dim and summary on "Hitters" databse from ISLR
#To find mean home runs (HmRun) of hitters
```{r}
mean(Hitters$HmRun)
```
#To find median years (years) of hitters
```{r}
median(Hitters$Years)
```
#The mean homeruns shows the average of homeruns per hitter regardless of years whereas the median years shows the number of years played in the middle of the dataset

```{r}
fetch_modes <-function(Hitters){
uniq_hitters <- unique(Hitters)
freq_values <-tabulate(match(Hitters,uniq_hitters))
modes_years <- uniq_hitters[freq_values == max(freq_values)]
return(modes_years)
}
mode_hits <- fetch_modes(Hitters$Hits)
mode_hits
```
```{r}
mode_salary <- fetch_modes(Hitters$Salary)
mode_salary                        
```
```{r}
mode_division <- fetch_modes(Hitters$Division)
mode_division
```
```{r}
head(Hitters)
```
#transformation of numerical variable
```{r}
Hitters$Hits <- log(Hitters$Hits)
head(Hitters)
```
#plot histogram of home runs (hmrun)
```{r}
ggplot(Hitters, aes(x = HmRun)) +
geom_histogram(binwidth = 5, fill = "orange", color = "blue") +
labs(title = " Home Runs across Hitters", x = "HmRun", y = "Count") +
theme_minimal()
```
#Histogram interpretation:  more hitters (slightly less than 100 total) hit 5 home runs than any number
#scatterplot of years to home runs.  As the home runs count increases, the volume of players reaching the higher home run number decreases
```{r}
ggplot(Hitters, aes(x = Years, y = HmRun)) +
geom_point(size = 1.5, color = "red") +
labs( title = "Scatter plot of Home Runs over Years", x = "Years of Playing", y = "Home Runs")
```
#Interpretation: years playing seems to increase the number of home runs for a player.  In the early years, most of the home run dats is less than 10 homruns for years 0-6.  After year 6, the number of home runs is distributed over a larger range with a higher mean and median than early years.  For players with 20+ yars, the number of homeruns decreases again.  
