---
title: "Assignment5"
author: "Michael Roberto"
date: "2025-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(cluster)
library(dplyr)
library(knitr)
```
```{r}
# Load data
cereals <- read.csv("Cereals.csv")
```
```{r}
# Remove rows with missing values
cereals_clean <- na.omit(cereals)
```
```{r}
# Select numeric columns for clustering
numeric_cols <- cereals_clean %>% select(calories, protein, fat, sodium, fiber, carbo, sugars, potass, vitamins, shelf, weight, cups, rating)
```
```{r}
# Normalize data
numeric_scaled <- scale(numeric_cols)
```
### Hierarchical clustering using Agnes
```{r}
agnes_single <- agnes(numeric_scaled, method = "single")
agnes_complete <- agnes(numeric_scaled, method = "complete")
agnes_average <- agnes(numeric_scaled, method = "average")
agnes_ward <- agnes(numeric_scaled, method = "ward")

coefficients <- c(single = agnes_single$ac, complete = agnes_complete$ac, average = agnes_average$ac, ward = agnes_ward$ac)
coefficients

best_method <- names(which.max(coefficients))
best_method
```
## Dendrogram for best method
```{r}
plot(agnes_ward, which.plots = 2, main = "Dendrogram - Ward's Method")
```
### Interpretation of Results
Ward's method typically produces compact, spherical clusters and often has the highest agglomerative coefficient, indicating strong clustering structure. Based on the dendrogram, cutting the tree at 4 clusters seems optimal because it finds a balance between interpretability and separation.

- **Cluster 1**: High fiber, low sugar cereals (e.g., bran-based cereals) – likely the healthiest.
- **Cluster 2**: Moderate sugar and fiber, balanced nutrients.
- **Cluster 3**: High sugar, low fiber cereals – less healthy, often marketed to children.
- **Cluster 4**: Mixed cereals with moderate ratings and nutrients.

Healthy cereals for schools should come from Cluster 1, as they have high fiber, low sugar, and good ratings.

## Determine number of clusters
```{r}
clusters <- cutree(as.hclust(agnes_ward), k = 4)
table(clusters)
```
## Summary table of healthy cereals (Cluster 1)
```{r}
healthy_cereals <- cereals_clean[clusters == 1, c("name", "calories", "protein", "fat", "fiber", "sugars", "rating")]
kable(healthy_cereals, caption = "Healthy Cereals in Cluster 1")
```
## Cluster stability check
```{r}
set.seed(123)
index <- sample(1:nrow(numeric_scaled), size = floor(0.5 * nrow(numeric_scaled)))
partA <- numeric_scaled[index, ]
partB <- numeric_scaled[-index, ]

agnes_A <- agnes(partA, method = "ward")
clusters_A <- cutree(as.hclust(agnes_A), k = 4)

centroids <- aggregate(partA, by = list(cluster = clusters_A), FUN = mean)
assign_to_centroid <- function(row, centroids) {
  dists <- apply(centroids[,-1], 1, function(centroid) sum((row - centroid)^2))
  return(which.min(dists))
}
assigned_B <- apply(partB, 1, assign_to_centroid, centroids = centroids)
assigned_B
```
## Healthy cereals cluster summary
```{r}
aggregate(numeric_scaled, by = list(cluster = clusters), FUN = mean)
```
### Should data be normalized?
Yes, normalization of data is important because variables are on different scales (such as sodium versus fiber). Without the normalization process, clustering would be dominated by variables with large ranges ultimately skewing statistical significance and results.