---
title: "Assignment 4"
author: "Michael Roberto"
date: "2025-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load required libraries
library(readr)        # For reading CSV files
library(dplyr)        # For data manipulation
library(cluster)      # For clustering algorithms
library(factoextra)   # For visualizing clustering results
```
```{r}
# Load the dataset from CSV file
p <- read_csv("Pharmaceuticals.csv")
```
```{r}
head(p)
```
```{r}
# Collect only the the quantitative variables (1-9) to cluster the 21 firms.
p1 <- p[,3:11]
head(p1)
```
```{r}
#Scale all of the quantitative variables in the dataframe.
ps <- scale(p1)
head(ps)
```
```{r}
#Determine the number of clusters to conduct the cluster analysis
wss <- (nrow(ps)-1)*sum(apply(ps,2,var))
wss
```
```{r}
for (i in 2:15) wss[i] <- sum(kmeans(ps,centers=i)$withinss)
wss
```
```{r}
plot(1:15, wss, type="b", xlab="Clusters",ylab="WSS - Within groups sum of squares")
```
#Analysis: the plot shows that 5 clusters will be sufficient to capture the majority of variations in the data
```{r}
#K-Means Cluster Analysis - this step fits the data with 5 clusters
fit <- kmeans(ps, 5)
#This step provides the mean value of all quantitative variables for each of the 5 clusters
aggregate(ps,by=list(fit$cluster),FUN=mean)
```
```{r}
ps1 <- data.frame(ps, fit$cluster)
ps1
```
```{r}
clusplot(ps, fit$cluster, color=TRUE, shade=TRUE, labels=2, lines=0)
```
# -----------------------------
# Analysis and Interpretation
# -----------------------------

#----Section B Analysis-------
#Cluster 1 represents rows 1, 4, 10 and 21
#Cluster 2 represents rows 2, 3, 5, 18 and 20
#Cluster 3 represents rows 7, 13, 15, 16 and 19
#Cluster 4 represents rows 6, 8, 9, 12 and 14
#Cluster 5 represents rows 11 and 17
#---Based on the aggregate command output, we can deduce the following from each cluster:
#Cluster 1 has the lowest Beta, revenue growth and asset turnover
#Cluster 2 has the highest PE ratio, but lowest ROE and profit margin
#Cluster 3 has the lowest leverage and PE ratio
#Cluster 4 has the lowest market cap and ROA with highest Beta and leverage
#Cluster 5 has the highest market cap, ROE, ROA, revenue growth, profit margin and asset turnover

#-----Section C Analysis-------
#Cluster 1 represented firms that included low revenue growth and asset turnover which aligned with two firms being hold and the #others being moderate sell and buy.  3 of the firms are based in the US.  
#Cluster 2 represented international firms that trended towards buy (3 firms) with two (US-based) being split between moderate sell and hold.  
#Cluster 3 represented US firms primarily that trended towards hold (3 firms) with two being split between moderate sell and hold.
#Cluster 4 represented US firms primarily that included a mixture of hold, buy and sell.
#Cluster 5 represented a mixture of US and UK firms primarily that included a mixture of hold and buy.

#----Section D Analysis-------
#Cluster 1: US-influenced stable and slow movers
#Cluster 2: International-influenced with growth expectations but not profitable yet
#Cluster 3: Low Leverage cluster
#Cluster 4: Small, volatile and debt-ridden
#Cluster 5: Large High Growth and Profit
