---
title: "k-NN Analysis in R"
author: "Michael Roberto"
date: "2025-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
getwd()
```
```{r}
# Load all necessary libraries for the assignment
library(class)  
library(caret)  
library(dplyr)   
library(ggplot2) 
library(readr)   

```
```{r}
# Load dataset for assignment and remove the two unnecessary columns
data <- read.csv("UniversalBank.csv") %>% select(-ID, -ZIP.Code)
summary(data)
```
```{r}
# Convert categorical variable (Education) into three dummy variables
data <- data %>%
  mutate(
    Education_1 = ifelse(Education == 1, 1, 0),
    Education_2 = ifelse(Education == 2, 1, 0),
    Education_3 = ifelse(Education == 3, 1, 0)
  ) %>%
  select(-Education)

```
```{r}
#Normalize the numerical predictions
normalize <- function(x) { (x - min(x)) / (max(x) - min(x)) }
data[, 1:7] <- as.data.frame(lapply(data[, 1:7], normalize))

```
```{r}
# Split data into training set at 60% and validation set at 40%
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.6, list = FALSE)
trainData <- data[trainIndex, ]
validData <- data[-trainIndex, ]

```
```{r}
# Define the predictor train/valid variables
train_X <- trainData %>% select(-Personal.Loan)
train_Y <- trainData$Personal.Loan
valid_X <- validData %>% select(-Personal.Loan)
valid_Y <- validData$Personal.Loan

```
```{r}
# Apply the k-NN where k = 1 to determine outcomes
knn_pred_k1 <- knn(train = train_X, test = valid_X, cl = train_Y, k = 1)
confusionMatrix(table(knn_pred_k1, valid_Y))

```
```{r}
# Find best k by evaluating accuracy for when k is between 1 and 20
k_values <- seq(1, 20, 2)
results <- data.frame(k = k_values, Accuracy = NA)

for (i in seq_along(k_values)) {
  knn_pred <- knn(train_X, valid_X, train_Y, k = k_values[i])
  results$Accuracy[i] <- sum(knn_pred == valid_Y) / length(valid_Y)
}
results
```
#Based on results above, K1 has the highest accuracy whoich balances overfitting and ignoring predictor information
```{r}
# Plot accuracy vs k
ggplot(results, aes(x = k, y = Accuracy)) +
  geom_line() + geom_point() +
  ggtitle("Best k") +
  xlab("Number of Neighbors (k)") +
  ylab("Accuracy")
```
```{r}
# Choose best k and evaluate performance
best_k <- results$k[which.max(results$Accuracy)]
knn_pred_best_k <- knn(train_X, valid_X, train_Y, k = best_k)
confusionMatrix(table(knn_pred_best_k, valid_Y))
```
```{r}
# Classify a new customer using the best k
new_customer <- data.frame(
  Age = normalize(40), Experience = normalize(10), Income = normalize(84),
  Family = normalize(2), CCAvg = normalize(2), Mortgage = normalize(0),
  Securities.Account = 0, CD.Account = 0, Online = 1, CreditCard = 1,
  Education_1 = 0, Education_2 = 1)
```
```{r}
# Repartition data between 50 training, 30 validation and 20 test
set.seed(123)
trainIndex <- createDataPartition(data$Personal.Loan, p = 0.5, list = FALSE)
trainData <- data[trainIndex, ]
remainingData <- data[-trainIndex, ]
validIndex <- createDataPartition(remainingData$Personal.Loan, p = 0.6, list = FALSE)
validData <- remainingData[validIndex, ]
testData <- remainingData[-validIndex, ]
```
```{r}
# Train and evaluate k-NN on new the 50/30/20 partitions
train_X <- trainData %>% select(-Personal.Loan)
train_Y <- trainData$Personal.Loan
valid_X <- validData %>% select(-Personal.Loan)
valid_Y <- validData$Personal.Loan
test_X <- testData %>% select(-Personal.Loan)
test_Y <- testData$Personal.Loan
```
```{r}
knn_test_pred <- knn(train_X, test_X, train_Y, k = best_k)
confusionMatrix(table(knn_test_pred, test_Y))
```
```{r}
#Comparison of confusion matrix between train/test/validation sets
knn_train_pred <- knn(train_X, train_X, train_Y, k = best_k)
train_conf_matrix <- confusionMatrix(table(knn_train_pred, train_Y))

knn_valid_pred <- knn(train_X, valid_X, train_Y, k = best_k)
valid_conf_matrix <- confusionMatrix(table(knn_valid_pred, valid_Y))

knn_test_pred <- knn(train_X, test_X, train_Y, k = best_k)
test_conf_matrix <- confusionMatrix(table(knn_test_pred, test_Y))
```
```{r}
#Print of confusion matrix to compare outcomes
print("Train Confusion Matrix:"); print(train_conf_matrix)
print("Validation Confusion Matrix:"); print(valid_conf_matrix)
print("Test Confusion Matrix:"); print(test_conf_matrix)
```
#Assessment of project data:  Test accuracy shows 'real' world performance but if test accuracy drops, the model will overfit.  Assuming validation and test accuract perform at similar levels, the model provides a good generalization of behavior.  Test accuracy is 0.96, Validation accuracy is 0.9573 and Train accuracy is 1.0.  The training set is most accurate with the validation set being least accurate.  
